### 2023-05-05 Manuscript/Draft1.Rmd
manuscript update

As see in Table **cite VAST setting table**, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with $77/80\approx 96\%$ of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate. **count how many w/ cov better than w/ out cov?** There is more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covariate response function (linear vs normally distributed etc). 
<!-- **For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information? Which covariate is best? Good question for future research** -->

As noted in our results, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates. However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response. Figure \@ref(fig:strata-plot) shows the strata excluded from sampling and Figure \@ref(fig:CovResponse) depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario. Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata). As a result, when projecting into those strata as seen in the lower portion of Figure \@ref(fig:CovResponse), Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.

Estimation methods that have large, inaccurate swings (stratified mean) can lead to changes in quotas that do not correspond to the true population trend, which could have a compounding effect (can lead to quotas that are too high/low given an incorrect assumption of increase/decrease in biomass). Our model has a constant assumed mortality that accounts for fishing and natural death and will not account for impacts of these decisions. VAST may provide more consistent biomass predictions(?)

>>>>

As see in Table **cite VAST setting table**, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73% had a lower absolute error compared to the corresponding estimate without covariates ($117/160\approx 73\%$), and $77/80\approx 96\%$ of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate. There is more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covariate response function (linear vs polynomial etc). 
<!-- **For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information? Which covariate is best? Good question for future research** -->

 Of the ~27% of VAST models where covariates did not improve the estimate, most provided a comparable error (for ex, 0.11 vs 0.13). The instances where including covariates provided a significant different in error took place scenarios that included either an increasing temperature and/or reduced sampling domain. More specifically, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates. However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response. Figure \@ref(fig:strata-plot) shows the strata excluded from sampling and Figure \@ref(fig:CovResponse) depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario. Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata). As a result, when projecting into those strata as seen in the lower portion of Figure \@ref(fig:CovResponse), Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.

Estimation methods that produce large variation between yearly estimates as displayed by the stratified mean can potentially lead to changes in catch limits that do not correspond to the true population trend, which could have a compounding effect. For example, a large, increasing biomass estimate when the population has actually decreased and is fairly low could potentially lead to a windfall catch limit that further reduces the total biomass available the following year. A second overestimate the following year could then have a detrimental impact by reducing the population even further. Our population model has assumed a constant total mortality that accounts for both fishing and natural death, and therefore will not account for impacts of such management decisions. This type of question can be best explored with a management strategy evaluation.

VAST may provide more consistent biomass predictions(?)

### 2023-05-05 Manuscript/Draft1.tex
manuscript update

As see in Table \textbf{cite VAST setting table}, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with \(77/80\approx 96\%\) of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate. \textbf{count how many w/ cov better than w/ out cov?} There is more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covariate response function (linear vs normally distributed etc).

As noted in our results, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates. However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response. Figure \ref{fig:strata-plot} shows the strata excluded from sampling and Figure \ref{fig:CovResponse} depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario. Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata). As a result, when projecting into those strata as seen in the lower portion of Figure \ref{fig:CovResponse}, Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.

Estimation methods that have large, inaccurate swings (stratified mean) can lead to changes in quotas that do not correspond to the true population trend, which could have a compounding effect (can lead to quotas that are too high/low given an incorrect assumption of increase/decrease in biomass). Our model has a constant assumed mortality that accounts for fishing and natural death and will not account for impacts of these decisions. VAST may provide more consistent biomass predictions(?)

>>>>

As see in Table \textbf{cite VAST setting table}, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73\% had a lower absolute error compared to the corresponding estimate without covariates (\(117/160\approx 73\%\)), and \(77/80\approx 96\%\) of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate. There is more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covariate response function (linear vs polynomial etc).

Of the \textasciitilde27\% of VAST models where covariates did not improve the estimate, most provided a comparable error (for ex, 0.11 vs 0.13). The instances where including covariates provided a significant different in error took place scenarios that included either an increasing temperature and/or reduced sampling domain. More specifically, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates. However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response. Figure \ref{fig:strata-plot} shows the strata excluded from sampling and Figure \ref{fig:CovResponse} depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario. Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata). As a result, when projecting into those strata as seen in the lower portion of Figure \ref{fig:CovResponse}, Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.

Estimation methods that produce large variation between yearly estimates as displayed by the stratified mean can potentially lead to changes in catch limits that do not correspond to the true population trend, which could have a compounding effect. For example, a large, increasing biomass estimate when the population has actually decreased and is fairly low could potentially lead to a windfall catch limit that further reduces the total biomass available the following year. A second overestimate the following year could then have a detrimental impact by reducing the population even further. Our population model has assumed a constant total mortality that accounts for both fishing and natural death, and therefore will not account for impacts of such management decisions. This type of question can be best explored with a management strategy evaluation.

VAST may provide more consistent biomass predictions(?)
